---
title: "Bobblehead Promotions"
author: "Bryan Clark"
date: "1/12/2019"
output: html_document
---

```{r setup, include = FALSE}
# customize output options 
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 6, fig.asp = 0.618, out.width = "70%", 
                      fig.align = "center")

# libraries used in report
library(knitr)
library(kableExtra)
library(tidyverse)
library(tidymodels)
library(glue)
library(lubridate)
library(hrbrthemes)
library(tidyquant)

# set theme for plots
theme_set(theme_tq(base_family = "Avenir"))

source("R/create_categorical_summary.R")
source("R/create_numeric_summary.R")
```

## Background

For the purpose of this analysis, let's say that we work in the Marketing or Business Analytics department as a data scientist for the New York Mets (or even a strategic consulting firm hired by them). I do not have a particular affinity for the Mets, but I have recently moved to NYC so I wanted to pick one of the New York teams.

The marketing group likes to use giveaways to drive fan engagement and satisfaction. A question has come up in a recent meeting regarding the effect of ballpark giveaways on attendance. The belief is that we see a boost in attendance during these giveaways and should consider doing them more often as a means to increase revenue. 

Dataset is from the [Marketing Science](http://www.informit.com/promotions/modeling-techniques-in-predictive-analytics-with-python-141177) book by Thomas Miller. 


## Executive Summary

While we do see an increase in attendance for games with promotions, when controlling for month and day of week, we do not see a statistically meaningful effect on attendance due to promotion. While this means that promotions may not necessarily drive an increase in attendance, we can say that it shows we did a good job at selecting games in which to have promotions since attendance was generally higher for those games. More fans receiving a promotion will have a better ROI on engagement for the promotion compared to if we had selected games with generally lower attendance. More data is needed to validate if our efforts led to an increasee in fan experience and engagement metrics. 

## Business Understanding

### Goals:

The ultimate goal of the analytics project is to determine if marketing giveaways have a positive effect on ballpark revenue. In addition to ticket sales, increasing attendance to the park can affect revenue streams from parking, merchandise, and concessions. It is possible to explore how different decisions made within these functions impact overall game-day revenue, but we will just focus on attendance. 

*[Attendance Prediction]*
A regression model could be built to determine if marketing promotions increase attendace, when controlling for other factors such as the day of week, oppoenent, team record, or weather. If marketing promotions have an impact on attendance, the model could be used to predict attendance for future games with or without a marketing promotion. Adding in associated costs, we could estimate the contribution margin of the promotion and determine which games would benefit most from the promotion. 

## Data Understanding

Each of these potential solutions assumes the data is available to move forward. To predict attendance for individual games, we will need data that consists of attendance for home games as well as features pertaining to the visiting team, weather, day/time, and promotions that ran. The hypothesis is that attendance is a function of the month, day of week, time of game, weather, recent performance of the home team, reputation of the visiting team, and any promotions that are running. 

Other factors that could impact attendance include if a particular pitcher is scheduled to play, winning/losing streaks the team has, or injuries that prevent star players from being in the lineup. In the case of the 2012 Mets, they had a pitcher win the CY Young award, so his starts could have boosted attendance ( [Wikipedia](https://en.wikipedia.org/wiki/2012_New_York_Mets_season)). 

Normally, we would query needed data from the database (likely using SQL). In this case, we are given a dataset that consists of some of the data we requested. It consists of home games and attendace for the 2012 season along with variables for month, day of week, the opposing team, some weather information, and whether one of three promotional giveaways ran (cap, shirt, or bobblehead). 

Starting in 2012, the capacity of Citi Field is [41,922](https://en.wikipedia.org/wiki/Citi_Field).

```{r data}
# load data
games_all_teams <- read_csv("data/bobbleheads.csv")
```

```{r}
# select team of interest
games <- games_all_teams %>%
  filter(home_team == "New York Mets")

games %>% glimpse()
```

At first glance, we see only 80 home games (opposed to a standard 81) for the New York Mets in 2012. [Baseball Reference](https://www.baseball-reference.com/teams/NYM/2012-schedule-scores.shtml) shows 81 home games, so we'd want to confirm if this is a data error or due to some other circumstance, such as a home game re-scheduled for another venue. 

We see games pretty evenly dispersed across month and day of week. Most of the games occurred with cloudy skies or at night. There were 17 total games with a promotional giveaway. 

Using the categorical and numeric summary tables, there were 17 total opponents with the most games played against teams within the division. The average temperature for home games was 73 degrees with a high of 94 and low of 49. Most importantly, the average attendace was 28k with a high that is slightly above capacity. This could be concerning as a data error, but sometimes attendance is slightly above capacity, so we won't raise the alarm bells and can confirm this finding with operations. 


```{r}
# view counts by categorical variable
games %>%
  select_if(is.character) %>%
  map(~ table(.))
```

```{r}
# view summary stats for categorical variables via custom function
categorical_summary_pre <- cat_summary_tbl(data = games)

categorical_summary_pre %>%
  mutate(variable = rownames(.)) %>%
  select(variable, everything())

# uncomment to display in HTML table
# kable(categorical_summary
      #,type = "html"
#      ) %>%
#  kable_styling(bootstrap_options = "striped", full_width = F, position = "left", 
#                latex_options = "scale_down")
```

```{r}
# view summary stats for numberical variables via custom function
numeric_summary_pre <- num_summary_tbl(data = games)

numeric_summary_pre %>%
  mutate(variable = rownames(.)) %>%
  select(variable, everything())

# uncomment to display in HTML table
# kable(numeric_summary
      #,type = "html"
#      ) %>%
#  kable_styling(bootstrap_options = "striped", full_width = F, position = "left", 
#                latex_options = "scale_down")
```

## Data Preparation

To prepare our data for modeling, we'll first convert our categorical variables to factors and then create a flag for any games in which a cap, shirt, or bobblehead was given away to fans. We then re-review our data to confirm the transformations were completed correctly. 

```{r}

# create factor levels and convert data
month_levels <- c(
  "APR", "MAY", "JUN", "JUL", "AUG", "SEP"
)

day_levels <- c(
  "Monday", "Tuesday", "Wednesday", "Thursday",
  "Friday", "Saturday", "Sunday"
)

games_tbl <- games %>%
  mutate_if(is.character, as.factor) %>%
  mutate(
    month = month %>% fct_relevel(month_levels),
    day_of_week = day_of_week %>% fct_relevel(day_levels),
    promotion = ifelse(cap == "YES" | shirt == "YES" | bobblehead == "YES", "YES", "NO"),
    promotion = promotion %>% as_factor()
    ) %>%
  select(month, day_of_week, opponent, skies, day_night, promotion, temp, attend)
```


```{r}
# review updated levels
games_tbl %>%
  select_if(is.factor) %>%
  map(levels)
```

```{r}
# confirm counts
games_tbl %>%
  select_if(is.factor) %>%
  map(~ table(.))
```


```{r}
categorical_summary <- cat_summary_tbl(data = games_tbl)

categorical_summary %>%
  mutate(variable = rownames(.)) %>%
  select(variable, everything())

# # uncomment to display in HTML table
# kable(categorical_summary
      #,type = "html"
#      ) %>%
#  kable_styling(bootstrap_options = "striped", full_width = F, position = "left", 
#                latex_options = "scale_down")
```

```{r}
numeric_summary <- num_summary_tbl(data = games_tbl)

numeric_summary %>%
  mutate(variable = rownames(.)) %>%
  select(variable, everything())

# # uncomment to display in HTML table
# kable(numeric_summary
      #,type = "html"
#      ) %>%
#  kable_styling(bootstrap_options = "striped", full_width = F, position = "left", 
#                latex_options = "scale_down")
```

Next we will plot our data to get a better understanding of the different distributions and relationships to attendance. 

__Attendance Histogram__
Attendance is mostly right-skewed with most values between 25-30k. There are a couple games with attendance over 40k.

```{r attendance_histogram}
games_tbl %>%
  ggplot(aes(x = attend / 1000)) +
  geom_histogram(bins = 30, color = "black", fill = "grey") +
  labs(
    title = "New York Mets Home Attendance (2012)",
    x = "Attendance (1000s)",
    y = "# of Games",
    caption = "Source: MLB League Data"
  )
```

__Promotion Boxplot__

We see a good initial indication in higher attendance associated with promotion. The boxplot for promotion vs. no promotion shows a slightly higher distribution for games with promotions. 

```{r promotion_boxplot}
games_tbl %>%
  ggplot(aes(x = promotion, y = attend / 1000, color = promotion)) +
  geom_boxplot() +
  scale_color_manual(values = c("grey", "blue")) +
  labs(
    title = "New York Mets Home Attendance (2012)",
    x = "",
    y = "Attendance (1000s)",
    caption = "Source: MLB League Data"
  ) +
  guides(
    color = FALSE
  )
```

__Month Boxplot__

Our boxplot for attendance across months shows attendance peaking in June and July with attendance dipping back down in August and September as the season winds down. 

```{r month_boxplot}
games_tbl %>% 
  ggplot(aes(x = month, y = attend / 1000)) +
  geom_boxplot() +
  labs(
    title = "New York Mets Home Attendance (2012)",
    x = "",
    y = "Attendance (1000s)",
    caption = "Source: MLB League Data"
  )
```

__Day of Week Boxplot__

Our boxplot for attendance across weekdays shows attendance generally higher on Saturdays with potential outliers on a few of the days. 

```{r day_boxplot}
games_tbl %>% 
  ggplot(aes(x = day_of_week, y = attend / 1000)) +
  geom_boxplot() +
   scale_x_discrete(
    labels = c("Mon","Tues","Wed", "Thurs", "Fri", "Sat", "Sun")
    ) +
  labs(
    title = "New York Mets Home Attendance (2012)",
    x = "",
    y = "Attendance (1000s)",
    caption = "Source: MLB League Data"
  )
```

__Day Time Boxplot__ 

Our boplot for day time of games shows higher attendance for day games compared to night games. 

```{r day_time_boxplot}
games_tbl %>% 
  ggplot(aes(x = day_night, y = attend / 1000)) +
  geom_boxplot() +
  labs(
    title = "New York Mets Home Attendance (2012)",
    x = "",
    y = "Attendance (1000s)",
    caption = "Source: MLB League Data"
  )
```

__Skies Boxplot__ 

Our boplot for the weather/skies of games shows higher attendance for games with clear skies compared to when it is cloudy. 

```{r skies_boxplot}
games_tbl %>% 
  ggplot(aes(x = skies, y = attend / 1000)) +
  geom_boxplot() +
  labs(
    title = "New York Mets Home Attendance (2012)",
    x = "",
    y = "Attendance (1000s)",
    caption = "Source: MLB League Data"
  )
```

__Day Time by Skies Boxplot__ 

Interestingly, when we facet the weather conditions by day and night games, we see night games have potentially higher attendance for cloudy skies with day games having better attendance for clear skies. This makes sense as a day baseball game is much more enjoyable in sunshine than in clouds, and the skies may matter less at night. 

```{r day_skies_boxplot}
games_tbl %>% 
  ggplot(aes(x = skies, y = attend / 1000)) +
  geom_boxplot() +
  labs(
    title = "New York Mets Home Attendance (2012)",
    x = "",
    y = "Attendance (1000s)",
    caption = "Source: MLB League Data"
  ) +
  facet_wrap(~ day_night, ncol = 2)
```

__Temperature and Attendance Scatter Plot__

We don't see a strong relationship between temperature and game attendance. When adding color for games with promotions, we do continue to see that games with promotions appear to have higher attendances.

Adding skies and time of game seem to not add any clarity to the relationship between temperature and attendance. 

```{r temp_attend_scatter}
games_tbl %>% 
  ggplot(aes(x = temp, y = attend / 1000, color = promotion)) +
  geom_point(size = 3, alpha = 1/2) +
  scale_color_manual(values = c("grey", "blue")) +
  labs(
    title = "New York Mets Home Attendance (2012)",
    x = "Temperature (F)",
    y = "Attendance (1000s)",
    caption = "Source: MLB League Data",
    color = "Promotion"
  )
```

```{r day_skies_facet, fig.asp = 1}
games_tbl %>% 
  ggplot(aes(x = temp, y = attend / 1000, color = promotion)) +
  geom_point(size = 3, alpha = 1/2) +
  scale_color_manual(values = c("grey", "blue")) +
  labs(
    title = "New York Mets Home Attendance (2012)",
    x = "Teemperature (F)",
    y = "Attendance (1000s)",
    color = "Promotion",
    caption = "Source: MLB League Data"
  ) +
  facet_grid(vars(skies), vars(day_night))
```

__Opponent Scatter__

When we look at attendance and promotions across different opponents, we see some oppoenents with no promotion (Atlanta Braves) and some where all games had a promotion (NY Yankees). Due to the size of our sample and lack of coverage, this may not be a good predictor to include in its current form. 

```{r opponent_scatter, fig.asp = 1}
games_tbl %>% 
  mutate(
    opponent = opponent %>%fct_rev()
  ) %>%
  ggplot(aes(x = opponent, y = attend / 1000, color = promotion)) +
  geom_point(size = 3, alpha = 1/2) +
  coord_flip() +
  scale_color_manual(values = c("grey", "blue")) +
  labs(
    title = "New York Mets Home Attendance (2012)",
    subtitle = "By Opponent",
    x = "",
    y = "Attendance (1000s)",
    caption = "Source: MLB League Data",
    color = "Promotion"
  )
```


## Modeling

Before moving forward with a regression model, we should first check to see if there is a statistically meaningful increase in attendance for games with a promotion. To do that, we can simulate a distribution where there is no difference, and then calculate the probability (p-value) we would see a difference as extreme or more extreme as the one we see with our sample. 

In doing this, we see a p-value of less than 2%, which is enough to assert there is a difference in attendance. However, based on our exploratory analysis, it is possible this difference could have occured due to other factors and we will want to verify this with a larger regression model. 

```{r}
# calculate difference between promotion and no promo
d_hat <- games_tbl %>% 
  group_by(promotion) %>% 
  summarize(mean_stat = mean(attend)) %>%
  summarize(diff(mean_stat)) %>%
  pull()

# use infer package to simulate null hypothesis of no difference
null_distn <- games_tbl %>%
  specify(response = attend, explanatory = promotion) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 10000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("YES", "NO"))

# plot null against observed
null_distn %>%
  ggplot(aes(x = stat)) +
  geom_histogram(bins = 30, color = "black", fill = "grey") +
  geom_vline(xintercept = d_hat, color = "red", linetype = "dashed") +
  annotate("text", x = d_hat + 1100, y = 500, label = "Observed\nDifference") +
  labs(
    title = "Promotion vs. No Promotion Null Hypothesis Simulation",
    subtitle = "10,000 trials",
    x = "Difference in Average Attendance (Promotion - No Promotion)",
    y = "Trials",
    caption = "Source: Simulated Data"
  )

```

```{r}
# calculate proportion of trials with a value as extreme or more
# multiple by two for two-tailed test
p_val <- null_distn %>%
  summarize(p_value = mean(stat >= d_hat) * 2) %>%
  pull()

p_val
```

As a simple first step, we will contruct a regression model using month, day of week, day time, and promotion as predictors. We can use the p-value to determine the probability of getting the coefficient (effect) given there really isn't an effect on our dependent variable. In this case, we'll want to look at both the p-value to determine statistical significance and coefficient to determine the size of the effect. Since we log-trnasform our attendance in our model, the coefficient will inform us of the percentage increase on attendance when a promotion runs. 

Unfortunately, we do not see a statistically meaninful signal being detected for promotions on attendane. 

```{r}
# log transform attendance and create regression model/summary
mod_form <- as.formula(log(attend) ~ month + day_of_week + day_night + promotion)
full_mod <- lm(mod_form, data = games_tbl)
summary(full_mod)
```

## Evaluation

The simulation test shows that there is a statistically meaningful increase in attendance for games with a promotion. Through further analysis, we could quantify the expected range of attendance increase, but this is unnecessary as the increase could have been due to confounding variables like the month or day of week. That is why we then use the multivariate regression to control for month, day of week, and day time of the games. When doing that, we no longer see a statistically detectable effect of promotions on attendance.

Why we see some predictive performance with an adjusted R-squared of .325, the variation in attendance is generally influenced by games taking place in June or September, games on Tuesday, Friday, or Saturday, and games at night versus games in the day time. 

We could explore additonal interactions, but our sample size is fairly small and our initial question has been answered. 

## Deployment

The takeaway from this is that we were not able to detect an increase in attendance for games with promotions compared to games without promotions, when controlling for month, day of week, and time of the games. 

That said, there was a statistically meaningful increase in attendance for games with promotions, so we did select good games to run a promotion to impact customer satisfaction and engagement. We would need additional data to validate if satisfaction was higher due to our efforts, but the positive takeaway is that productive games were chosen to run promotions. If we ran promotions on games with a lower on average attendance, then we would get a reduced potential ROI on our engagement investment as it would have been made on a smaller number of fans. 
